---
title: "Statial"
date: "`r BiocStyle::doc_date()`"
params:
  test: FALSE
author:
- name: Farhan Ameen
  affiliation:  
  - School of Mathematics and Statistics, University of Sydney, Australia
- name: Sourish Iyengar
  affiliation:
  - School of Mathematics and Statistics, University of Sydney, Australia
- name: Ellis Patrick
  affiliation:
  - &WIMR Westmead Institute for Medical Research, University of Sydney, Australia
  - School of Mathematics and Statistics, University of Sydney, Australia
vignette: >
  %\VignetteIndexEntry{"Introduction to Statial"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(BiocStyle)
```


```{r warning = FALSE, message = FALSE}
# Loading required packages
library(Statial)
library(spicyR)
library(ClassifyR)
library(tidyverse)
library(SingleCellExperiment)
library(ggplot2)

theme_set(theme_classic())
```


# Installation
```{r eval = FALSE}
# Install the package from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("Statial")
```

# Overview

There are over 37 trillion cells in the human body, each taking up different forms and functions. The behaviour of these cells can be described by canonical characteristics, but their functions can also dynamically change based on their environmental context, leading to cells with diverse states. Understanding changes in cell state and the interplay between cells is key to understanding their mechanisms of action and how they contribute to human disease. `Statial` is a suite of functions for identifying changes in cell state. This guide will provide a step-by-step overview of some key functions within `Statial`. 


# Evaluating cell localisation 

`Kontextual` is a method to evaluate the localisation relationship between two cell types in an image. `Kontextual` builds on the L-function by contextualising the relationship between two cell types in reference to the typical spatial behaviour of a  $3^{rd}$ cell type/population. By taking this approach, `Kontextual` is invariant to changes in the window of the image as well as tissue structures which may be present. 

The definitions of cell types and cell states are somewhat ambiguous, cell types imply well defined groups of cells that serve different roles from one another, on the other hand cell states imply that cells are a dynamic entity which cannot be discretised, and thus exist in a continuum. For the purposes of using `Kontextual` we treat cell states as identified clusters of cells, where larger clusters represent a "parent" cell population, and finer sub-clusters representing a "child" cell population. For example a CD4 T cell may be considered a child to a larger parent population of Immune cells. `Kontextual` thus aims to see how a child population of cells deviate from the spatial behaviour of their parent population, and how that influences the localisation between the child cell state and another cell state.

# Loading example data

In the following we will analyse cutaneous squamous cell carcinoma data from the head and neck region (Ferguson et al. 2022). These images are stored in a `SingleCellExperiment` object. The first step in analysing these images is to organise all the cells present into cell state populations or clusters, e.g. all the different B cell types are put in a vector called bcells.

```{r}
# Load head and neck data
data("kerenSCE")

# Examine all cell types in image
unique(kerenSCE$cellType)

# Set up cell populations
tumour <- c("Keratin_Tumour", "Tumour")

bcells <- c("B")
tcells <- c("CD3_Cell", "CD4_Cell", "CD8_Cell", "Tregs")
myeloid <- c("Dc/Mono", "DC", "Mono/Neu", "Macrophages", "other immune", "Neutrophils")

endothelial <- c("Endothelial")
mesenchymal <- c("Mesenchymal")

tissue <- c(endothelial, mesenchymal)
immune <- c(bcells, tcells, myeloid, "NK") # GC = granulocytes

all <- c(tumour, tissue, immune, "Unidentified")
```


# Kontextual

`Kontextual` accepts a `SingleCellExperiment` object, or a single image, or list of images from a `SingleCellExperiment` object, this gets passed into the `cells` argument. The two cell types which will be evaluated are specified in the `to` and `from` arguments. A parent population must also be specified in the `parent` argument, note the parent cell population must include the `to` cell type.  The argument `r` will specify the radius which the cell relationship will be evaluated on. `Kontextual` supports parallel processing, the number of cores can be specified using the `cores` argument. `Kontextual` can take a single value or multiple values for each argument and will test all combinations of the arguments specified. 


Here we test the relationship between CD4 T cells (`TC_CD4`) and one type of Tumour cell (`SC1`) with all immune cells being our parent population. The output is a data frame with the `original` L-function values and `kontextual` values as well as the values for the arguments specified. Where a negative `original` or `kontextual` value represents dispersion between `TC_CD4` and `SC1` and a positive value indicates localisation. 

```{r}
CD4_Kontextual <- Kontextual(
  cells = kerenSCE,
  r = 50,
  from = "CD4_Cell",
  to = "Keratin_Tumour",
  parent = immune,
  cores = 1
)

head(CD4_Kontextual)
```

The `original` and `kontextual` values can be compared in a scatter plot to identify relationships which have been identified differently between the functions. The dashed red line represents the expected pattern of the data, i.e. where there exists no relationship between the two cell types. The top left quadrant therefore represents values where `Kontextual` has identified localisation between `CD4` and `SC1` and the original L-function has identified dispersion, the opposite is true for the bottom right quadrant.

```{r fig.wide = TRUE}

ggplot(CD4_Kontextual, aes(x = original, y = kontextual, col = imageID)) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, col = "red", linetype = "dashed") +
  geom_vline(xintercept = 0, col = "red", linetype = "dashed") +
  theme(
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  )
```

Alternatively all pairwise cell relationships and their corresponding parent in the dataset can be tested. A data frame with all pairwise combinations can be creating using the `parentCombinations` function. This function takes in a vector of all the cells, as well as all the parent vectors set up earlier. As shown below the output is a data frame specifying the `to`, `from`, and `parent` arguments for `Kontextual`.

```{r}
# Get all relationships between cell types and their parents
parentDf <- parentCombinations(
  all = all,
  tumour,
  bcells,
  tcells,
  myeloid,
  endothelial,
  epithelial,
  tissue,
  immune
)
head(parentDf)
```


Rather than specifying `to`, `from`, and `parent` in Kontextual, the output from `parentCombinations` can be inputed into `Kontextual` using the `parentDf` argument. Here we will test the first 20 pairwise cell relationships in `parentCombinations` on the first image of the dataset.
```{r}
# Selecting Image 1 as an example
image_1 <- headSCE |>
  colData() |>
  data.frame() |>
  filter(imageID == "1")


image1_Kontextual <- Kontextual(
  cells = image_1,
  parentDf = parentDf[1:20, ],
  r = 50,
  cores = 1
)

head(image1_Kontextual)
```


# RsCurve

Here we examine an image were the relationship between 2 cell types depends on a parent cell population. This is a breast cancer image obtained from Keren et al, 2018 from patient 6. In the image below we can see that `p53` and `Immune` are dispersed. However when the behaviour of `p53` is placed in the context of the spatial behaviour of its parent population `Keratin+Tumour`, `p53` and `Immune` now appear localised.


```{r}
data("kerenImage")

kerenImage |>
  filter(cellType %in% c("Keratin+Tumour", "Immune", "p53")) |>
  arrange(cellType) |>
  ggplot(aes(x = x, y = y, color = cellType)) +
  geom_point(size = 1) +
  scale_colour_manual(values = c("#505050", "#D6D6D6", "#64BC46"))
```


The `rsCurve` function plots the L-function value and Kontextual values over a range of radii. If the points lie above the red line (expected pattern) then localisation is indicated for that radius, if the points lie below the red line then dispersion is indicated. As seen in the following plot Kontextual is able to correctly identify localisation between `p53` and `Immune` in the example image for a certain range of radii. When the radius gets too large the overall relationship between `p53` and `Immune` looks dispersed. The original L-function is not able to identify localisation at any value of radii.

```{r}
rsDf <- rsCurve(
  cells = kerenImage,
  from = "p53",
  to = "Immune",
  parent = c("p53", "Keratin+Tumour"),
  rs = seq(10, 510, 100),
  cores = 1
)


ggplotRs(rsDf)
```


##Sourish's Stuff

```{r}
data(kerenSCE)

SCE <- getDistances(kerenSCE,
                    Rs = c(200),
                    nCores = 30)

SCE <- getAbundances(SCE,
                     Rs = c(200),
                     nCores = 30
                     )

```

First, let's examine the effect of the distance between Keratin/Tumour cells and
macrophages on the expression of p53 in Keratin/Tumour cells. We can observe
that keratin/tumour cells closer to a group of macrophages tend to have higher
expression of p53. This relationship is outlined with the second graph, showing
a decrease of p53 expression in keratin/tumour cells over distance to 
macrophages. The pvalue provided also shows us that this is a significant 
interaction.
```{r}
visualiseImageRelationship(
  data = SCE,
  Rs = c(200),
  imageID = "3",
  mainCellType = "B",
  interactingCellType = "CD8_Cell",
  depedentMarker = "CD8",
  interactive = FALSE,
  plotModelFit = FALSE,
  method = "rlm",
  modelType = "dist200_"
)

```


Now, let's look at the most significant interactions across all the images.
```{r}
stateChanges <- getStateChanges(
  singleCellData = SCE,
  Rs = c(200),
  typeAll = c("dist200", "abundance200"),
  method = "lm",
  isMixed = FALSE,
  nCores = 40)

```

Let's take a look at the most significant gene-to-distance pairwise 
interactions. We can see that the most significant interaction between distance
and expression occurs between B cells and CD8+ T cells for the CD8 ligand.
CD8 is a highly specific ligand for CD8+ T cells, and should not be expressed 
on B cells in a normal biological context. 

```{r}
head(stateChanges)
```
Taking a deeper look at this interaction, we can indeed see that the result 
appears real. So why are B cells expressing CD8?

This brings us to a key problem of cell segmentation - contamination.
```{r}
visualiseImageRelationship(
  data = SCE,
  Rs = c(200),
  imageID = "3",
  mainCellType = "B",
  interactingCellType = "CD8_Cell",
  depedentMarker = "CD8",
  interactive = FALSE,
  plotModelFit = FALSE,
  method = "rlm",
  modelType = "dist200_"
)

```

Contamination
### Write explanation of contamination here

To circumvent this problem, Statial provides a function that predicts the
probability that a cell is any particular cell type. This comes in the format of
"rfMainCellProb". "rfMainCellProb" provides the probability that a cell is what
it's stated to be. E.g. For a cell designated as CD8, rfMainCellProb could give 
a 80% chance that the cell is indeed CD8, due to contamination. We can then
introduce these probabilities as covariates into our linear model.
```{r}
SCE <- calcContamination(SCE)

stateChangesContam <- getStateChanges(
  singleCellData = SCE,
  Rs = c(200),
  typeAll = c("dist200", "abundance200"),
  covariates = c("rfMainCellProb"),
  method = "lm",
  isMixed = TRUE,
  nCores = 30)
```


We can see that adding the contamination scores as covariates improves the...
TODO: generate the ROC plots (sorry could not figure this one out + could not
find code for this)
```{r}

```


The next steps will involve comparing survival time to see if these 
distance-to-gene interactions are biologically significant.

### Ellis, I've just realised that we're looking at survival data, not a
### discrete variable. Not sure if colTest is the correct choice (correct me if I'm wrong)

```{r}
stateChanges <- getStateChanges(
  singleCellData = SCE,
  Rs = c(200),
  typeAll = c("dist200", "abundance200"),
  method = "lm",
  isMixed = TRUE,
  nCores = 30)

modellingData <- listImageModelsCVFormat(stateChanges,
                                         values_from = "tValue",
                                         removeColsThresh = 0.2,
                                         missingReplacement = 0)
```



# References

Ferguson, A. L., Sharman, A. R., Allen, R. O., Ye, T., Lee, J. H., Low, T.-H. H., Ch'ng, S., Palme, C. E., Ashford, B., Ranson, M., Clark, J. R., Patrick, E., Gupta, R., & Palendira, U. (2022). High-Dimensional and Spatial Analysis Reveals Immune Landscape–Dependent Progression in Cutaneous Squamous Cell Carcinoma. Clinical Cancer Research, OF1-OF12. ([DOI](https://doi.org/10.1158/1078-0432.Ccr-22-1332))

Keren, L., Bosse, M., Marquez, D., Angoshtari, R., Jain, S., Varma, S., Yang, S. R., Kurian, A., Van Valen, D., West, R., Bendall, S. C., & Angelo, M. (2018). A Structured Tumor-Immune Microenvironment in Triple Negative Breast Cancer Revealed by Multiplexed Ion Beam Imaging. Cell, 174(6), 1373-1387.e1319. ([DOI](https://doi.org/10.1016/j.cell.2018.08.039))



# sessionInfo
```{r}
sessionInfo()
```
