---
title: "Statial"
date: "`r BiocStyle::doc_date()`"
params:
  test: FALSE
author:
- name: Farhan Ameen
  affiliation:  
  - School of Mathematics and Statistics, University of Sydney, Australia
- name: Sourish Iyengar
  affiliation:
  - School of Mathematics and Statistics, University of Sydney, Australia
- name: Ellis Patrick
  affiliation:
  - &WIMR Westmead Institute for Medical Research, University of Sydney, Australia
  - School of Mathematics and Statistics, University of Sydney, Australia
vignette: >
  %\VignetteIndexEntry{"Introduction to Statial"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(BiocStyle)
```


```{r warning = FALSE, message = FALSE}
# Loading required packages
library(Statial)
library(spicyR)
library(ClassifyR)
library(tidyverse)
library(SingleCellExperiment)
library(ggplot2)
library(ggsurvfit)
library(survival)

theme_set(theme_classic())
```


# Installation
```{r eval = FALSE}
# Install the package from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("Statial")
```

# Overview

There are over 37 trillion cells in the human body, each taking up different forms and functions. The behaviour of these cells can be described by canonical characteristics, but their functions can also dynamically change based on their environmental context, leading to cells with diverse states. Understanding changes in cell state and the interplay between cells is key to understanding their mechanisms of action and how they contribute to human disease. `Statial` is a suite of functions for identifying changes in cell state. This guide will provide a step-by-step overview of some key functions within `Statial`. 


# Evaluating cell localisation 

`Kontextual` is a method to evaluate the localisation relationship between two cell types in an image. `Kontextual` builds on the L-function by contextualising the relationship between two cell types in reference to the typical spatial behaviour of a  $3^{rd}$ cell type/population. By taking this approach, `Kontextual` is invariant to changes in the window of the image as well as tissue structures which may be present. 

The definitions of cell types and cell states are somewhat ambiguous, cell types imply well defined groups of cells that serve different roles from one another, on the other hand cell states imply that cells are a dynamic entity which cannot be discretised, and thus exist in a continuum. For the purposes of using `Kontextual` we treat cell states as identified clusters of cells, where larger clusters represent a "parent" cell population, and finer sub-clusters representing a "child" cell population. For example a CD4 T cell may be considered a child to a larger parent population of Immune cells. `Kontextual` thus aims to see how a child population of cells deviate from the spatial behaviour of their parent population, and how that influences the localisation between the child cell state and another cell state.

# Loading example data

In the following we will analyse breast cancer data from Keren et al. 2018. These images are stored in a `SingleCellExperiment` object. The first step in analysing these images is to organise all the cells present into cell state populations or clusters, e.g. all the different B cell types are put in a vector called bcells.

```{r}
# Load head and neck data
data("kerenSCE")

# Examine all cell types in image
unique(kerenSCE$cellType)

# Set up cell populations
tumour <- c("Keratin_Tumour", "Tumour")

bcells <- c("B")
tcells <- c("CD3_Cell", "CD4_Cell", "CD8_Cell", "Tregs")
myeloid <- c("Dc/Mono", "DC", "Mono/Neu", "Macrophages", "other immune", "Neutrophils")

endothelial <- c("Endothelial")
mesenchymal <- c("Mesenchymal")

tissue <- c(endothelial, mesenchymal)
immune <- c(bcells, tcells, myeloid, "NK") # GC = granulocytes

all <- c(tumour, tissue, immune, "Unidentified")
```


# Kontextual

`Kontextual` accepts a `SingleCellExperiment` object, or a single image, or list of images from a `SingleCellExperiment` object, this gets passed into the `cells` argument. The two cell types which will be evaluated are specified in the `to` and `from` arguments. A parent population must also be specified in the `parent` argument, note the parent cell population must include the `to` cell type.  The argument `r` will specify the radius which the cell relationship will be evaluated on. `Kontextual` supports parallel processing, the number of cores can be specified using the `cores` argument. `Kontextual` can take a single value or multiple values for each argument and will test all combinations of the arguments specified. 


Here we test the relationship between CD4 T cells (`TC_CD4`) and one type of Tumour cell (`SC1`) with all immune cells being our parent population. The output is a data frame with the `original` L-function values and `kontextual` values as well as the values for the arguments specified. Where a negative `original` or `kontextual` value represents dispersion between `TC_CD4` and `SC1` and a positive value indicates localisation. 

```{r}
CD4_Kontextual <- Kontextual(
  cells = kerenSCE,
  r = 50,
  from = "CD4_Cell",
  to = "Keratin_Tumour",
  parent = immune,
  cores = 1
)

head(CD4_Kontextual)
```

The `original` and `kontextual` values can be compared in a scatter plot to identify relationships which have been identified differently between the functions. The dashed red line represents the expected pattern of the data, i.e. where there exists no relationship between the two cell types. The top left quadrant therefore represents values where `Kontextual` has identified localisation between `CD4` and `SC1` and the original L-function has identified dispersion, the opposite is true for the bottom right quadrant.

```{r fig.wide = TRUE}

ggplot(CD4_Kontextual, aes(x = original, y = kontextual, col = imageID)) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, col = "red", linetype = "dashed") +
  geom_vline(xintercept = 0, col = "red", linetype = "dashed") +
  theme(
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  )
```

Alternatively all pairwise cell relationships and their corresponding parent in the dataset can be tested. A data frame with all pairwise combinations can be creating using the `parentCombinations` function. This function takes in a vector of all the cells, as well as all the parent vectors set up earlier. As shown below the output is a data frame specifying the `to`, `from`, and `parent` arguments for `Kontextual`.

```{r}
# Get all relationships between cell types and their parents
parentDf <- parentCombinations(
  all = all,
  tumour,
  bcells,
  tcells,
  myeloid,
  endothelial,
  mesenchymal,
  tissue,
  immune
)
head(parentDf)
```


Rather than specifying `to`, `from`, and `parent` in Kontextual, the output from `parentCombinations` can be inputed into `Kontextual` using the `parentDf` argument, to examine all pairwise relationships in the dataset. This chunk will take a signficant amount of time to run, for demonstration the results have been saved and are loaded in.


```{r eval = FALSE, show = FALSE}
# save(kerenKontextual, file = "../data/kerenKontextual.rda")
# tools::resaveRdaFiles("../data/kerenKontextual.rda", compress = "auto")
```

```{r eval = FALSE}
# Running Kontextual on all relationships across all images.
kerenKontextual <- Kontextual(
  cells = kerenSCE,
  parentDf = parentDf,
  r = 50,
  cores = 1
)

```


```{r}
data("kerenKontextual")

head(kerenKontextual)
```


# RsCurve

Here we examine an image from the Keren et al. 2018 dataset where the relationship between 2 cell types depends on a parent cell population. In the image below we can see that `p53` and `Immune` are dispersed. However when the behaviour of `p53` is placed in the context of the spatial behaviour of its parent population `Keratin+Tumour`, `p53` and `Immune` now appear localised.


```{r}
#Select image 6 from the kerenSCE dataset
kerenImage6 = kerenSCE[, kerenSCE$imageID =="6"]

#Select for all cells that express higher than baseline level of p53
p53Pos = assay(kerenImage6)["p53",]  |> 
  as.numeric() > -0.300460
kerenImage6$cellType[p53Pos & kerenImage6$cellType %in% c("Keratin_Tumour")] <- "p53+Tumour"

#Group all immune cells under the name "Immune"
kerenImage6$cellType[kerenImage6$cellType %in% immune] <- "Immune"

kerenImage6 |>
  colData() %>%
  as.data.frame() %>%
  filter(cellType %in% c("Keratin_Tumour", "Immune", "p53+Tumour")) %>%
  arrange(cellType) %>%
  ggplot(aes(x = x, y = y, color = cellType)) +
  geom_point(size = 1) +
  scale_colour_manual(values = c("#505050", "#D6D6D6", "#64BC46"))

```


The `rsCurve` function plots the L-function value and Kontextual values over a range of radii. If the points lie above the red line (expected pattern) then localisation is indicated for that radius, if the points lie below the red line then dispersion is indicated. As seen in the following plot Kontextual is able to correctly identify localisation between `p53` and `Immune` in the example image for a certain range of radii. When the radius gets too large the overall relationship between `p53` and `Immune` looks dispersed. The original L-function is not able to identify localisation at any value of radii.

```{r}
rsDf <- rsCurve(
  cells = kerenImage6,
  from = "p53+Tumour",
  to = "Immune",
  parent = c("p53", "Keratin_Tumour"),
  rs = seq(10, 510, 100),
  cores = 1
)

ggplotRs(rsDf)
```

# Examining Cell-to-cell interactions with marker expression
In the next section of this vignette, we will utilise multiplexed imaging data to computationally identify and quantify evidence of cell interactions that catalyse cell state changes. This approach measures how protein markers in a cell change with spatial proximity and abundance to other cell types. 
The methods utilised here will provide a framework to explore how the dynamic behaviour of cells are altered by the agents they are surrounded by.

The first step in analysing these changes is to calculate the spatial proximity (`getDistances`) and abundance (`getAbundances`) of each cell to every cell type. These values will then be stored in the `reducedDims` slot of the `SingleCellExperiment` object under the names `distances` and `abundances` respectively.
```{r}
data(kerenSCE)

SCE <- getDistances(kerenSCE,
                    Rs = c(200),
                    nCores = 30)

SCE <- getAbundances(SCE,
                     Rs = c(200),
                     nCores = 30
                     )

```

First, let's examine the same effect observed earlier with Kontextual - the distance between p53-positive keratin/tumour cells and macrophages in the context of total keratin/tumour cells. 
Statial provides a convenient function for observing this interaction - `visualiseImageRelationship`. Here, we can specify the `imageID` and our main cell of interest, `mainCellType` and our interacting cell type `interactingCellType` and our marker of interest `dependentMarker`, in this case p53.

Through this analysis, we can observe that keratin/tumour cells closer to a group of macrophages tend to have higher expression of p53, as observed in the first graph. This relationship is outlined with the second graph, showing an overall decrease of p53 expression in keratin/tumour cells over distance to macrophages. The pvalue provided also shows us that this correlative interaction is statistically significant.

These results are similar to those provided with Kontextual, which calculates a similar localisation between p53+ keratin/tumour cells and macrophages.
```{r}
visualiseImageRelationship(
  data = SCE,
  Rs = c(200),
  imageID = "6",
  mainCellType = "Keratin_Tumour",
  interactingCellType = "Macrophages",
  depedentMarker = "p53",
  interactive = FALSE,
  plotModelFit = FALSE,
  method = "lm",
  modelType = "dist200_"
)

```
Beyond looking at single cell-to-cell interactions for a single image, we can also look at all interactions across all images. Statial provides the `getStateChanges` function for this exact purpose, allowing us to examine the most significant correlations between distance and marker expression.
```{r}
stateChanges <- getStateChanges(
  singleCellData = SCE,
  Rs = c(200),
  typeAll = c("dist200", "abundance200"),
  method = "lm",
  isMixed = FALSE,
  nCores = 40)

head(stateChanges, 10)
```

Let's take a look at the top 10 most significant gene-to-distance pairwise interactions. We can see that the most significant interaction involving keratin/tumour cells occurs between keratin/tumour cells and macrophages in image 27, where HLA Class I molecules in keratin/tumour cells are upregulated when nearby macrophage populations. 
We can examine this interaction further with the `visualiseImageRelationship` function, showing us a clear visual correlation between macrophage density and HLA Class I expression.

Biologically, HLA Class I is a ligand which exists on all nucleated cells, tasked with presenting internal cell antigens for recognition by the immune system, marking aberrant cells for destruction by either CD8+ T cells or NK cells. This increased expression of HLA Class I molecules on tumour cells is an interesting finding, and potentially uncovers a novel interaction between tumour cells and macrophages which drives aberrant HLA Class I expression.

```{r}
visualiseImageRelationship(
  data = SCE,
  Rs = c(200),
  imageID = "27",
  mainCellType = "Keratin_Tumour",
  interactingCellType = "Macrophages",
  depedentMarker = "HLA_Class_1",
  interactive = FALSE,
  plotModelFit = FALSE,
  method = "lm",
  modelType = "dist200_"
)
```
However, these significant results are occasionally isolated to a single image or patient, and it may sometimes be wiser to perform a more holistic analysis of all the images/patients to examine common interactions across a whole group. Additionally, patients can be divided into treatment groups to examine the effects of a treatment on cell-to-cell interactions. 

Fortunately, `getStateChanges` provides the `isMixed` parameter for performing a holistic analysis. In this next section, we will employ a mixed effects model to examine the most significant cell-to-cell interactions shared across all images.


```{r}
stateChangesMixed <- getStateChanges(
  singleCellData = SCE,
  Rs = c(200),
  typeAll = c("dist200", "abundance200"),
  method = "lm",
  isMixed = TRUE,
  nCores = 40)

head(stateChangesMixed, 10)
```

Looking at the same top 10 most significant interactions, we can see that the results are vastly different. Running the mixed effects model appears to bring about some curious results. One of the most significant interactions occurs between B cells and CD8 T cells, where B cells are found to increase in CD8 expression when in close proximity to CD8 T cells. 
Biologically, CD8 is a highly specific ligand for CD8 T cells, and under no biologically health circumstances are they expressed in B cells. 

Could this potentially be an artefact of `getStateChanges`? Taking a look at an image through the `visualiseImageRelationships` function, we indeed observe an apparent localisation between CD8+ B cells and CD8 T cells.

QUESTIONBOX: Feel free to play around with the imageID parameter to observe the effect in other imageIDs.

```{r}
visualiseImageRelationship(
  data = SCE,
  Rs = c(200),
  imageID = "3",
  mainCellType = "B",
  interactingCellType = "CD8_Cell",
  depedentMarker = "CD8",
  interactive = FALSE,
  plotModelFit = FALSE,
  method = "lm",
  modelType = "dist200_"
)
```
So why are B cells expressing CD8?
This brings us to a key problem of cell segmentation - contamination.

# Contamination
Contamination, or lateral marker spill over is an issue that results in a cell’s marker expressions being wrongly attributed to another adjacent cell. This issue arises from incorrect segmentation 
where components of one cell are wrongly determined as belonging to another cell. 
Alternatively, this issue can arise when antibodies used to tag and measure marker expressions don't latch on properly to a cell of interest, thereby resulting in residual markers being wrongly assigned as belonging to a cell near the intended target cell. It is important that we either correct or account
for this incorrect attribution of markers in our modelling process. This is critical in 
understanding whether significant cell-cell interactions detected are an artefact of technical 
measurement errors driven by spill over or are real biological changes that represent a shift in a 
cell’s state.

To circumvent this problem, Statial provides a function that predicts the
probability that a cell is any particular cell type. This comes in the format of the `calcContamination` function, which returns a dataframe of probabilities demarcating the chance of a cell being any particular cellType. It also provides the `rfMainCellProb` vector, which provides the probability that a cell is indeed the cell type it has been designated. 
E.g. For a cell designated as CD8, rfMainCellProb could give a 80% chance that the cell is indeed CD8, due to contamination. We can then introduce these probabilities as covariates into our linear model.
```{r}
SCE <- calcContamination(SCE)

stateChangesContam <- getStateChanges(
  singleCellData = SCE,
  Rs = c(200),
  typeAll = c("dist200", "abundance200"),
  covariates = c("rfMainCellProb"),
  method = "lm",
  isMixed = TRUE,
  nCores = 30)

head(stateChangesContam)
```

We can see that adding the contamination scores as covariates improves the...
TODO: generate the ROC plots (sorry could not figure this one out + could not
find code for this)
```{r}
#TODO: ROC Plots + colTest(coxPH)
```


The next steps will involve comparing survival time to see if these 
distance-to-gene interactions are biologically significant.



# Relating `Statial` features to patient groups or outcomes.

To examine whether the features obtained from `Statial` are associated with patient outcomes or groupings, we can use the `colTest` function from `SpicyR`. To understand if features differ significantly between 2 patient groups, specify `type = "ttest"` in `colTest`. Here we examine which features are most associated with patient survival using the Kontextual values as example. To do so, the survival data is extracted from `kerenSCE` and converted to a survival object. In addition to this, the Kontextual results must be converted from a `data.frame` to a wide `matrix`, this can be done using `prepKontextMat`. Note, to extract the original L-function values, specify `type = "original"` in `prepKontextMat`. Finally, both the Kontextual matrix and survival object are passed into `colTest`, with `type = "coxph"` to obtain the survival results. 

As we can see from the results `Mesenchymal_Macrophages_tissue` is the most significant pairwise relationship which contributes to patient survival. That is the relationship between Mesenchymal cells and Macrophages, relative to the parent population of all tissue cells. We can see that there is a negative coeffcient associated with this relationship, which tells us the increase in localisation of Mesenchymal and Macrophages lead to more favourable survival outcomes for patients.  


```{r}
# Extracting survial data
survData = kerenSCE@colData |> 
    data.frame() |> 
    select(imageID, Survival_days_capped, Censored) |> 
    unique()

# Creating survival vector
kerenSurv = Surv(survData$Survival_days_capped, survData$Censored)
names(kerenSurv) = survData$imageID


# Converting Kontextual result into data matrix
kontextMat = prepKontextMat(kerenKontextual)

# Ensuring rownames of kontextMat match up with rownames of the survival vector 
kontextMat = kontextMat[match(names(kerenSurv), rownames(kontextMat)), ]


# Running survival analysis

coxTests = function(measurementMat, Surv) {
  
  result = apply(measurementMat, 2, function(measurementCol) {
    fit = coxph(Surv ~ measurementCol)
    summary(fit)$coefficients[1, c(1, 3, 5)]
  })
  
  result = result |> 
    t() |> 
    data.frame() |> 
    rownames_to_column() 
  
  colnames(result) = c("imageID", "coef", "se.coef", "p.value")
  result$`adj.p.value` = result$`p.value`*nrow(result)
  
  result = result |> arrange(p.value)
  
  return(result)
}

survialResults = coxTests(kontextMat, kerenSurv)

head(survialResults)

```
The association between `Mesenchymal_Macrophages_tissue` and survival can also be visualised on a Kaplan-Meier curve. We must first extract the Kontextual values of this relationship across all images. Next we determine if Mesenchymal and Macrophages are relatively attracted or avoiding in each image, by comparing the Kontextual value in each image to the median Kontextual value. Finally we plot the Kaplan-Meier curve using the `ggsurvfit` package. 

As shown below, when Mesenchymal and Macrophages are relatively more localised to one another, patients tend to have better survival outcomes. When these cells are relatively dispersed, patients have a poorer survival outcome. 

```{r}
# Selecting most significant relationship
survRelationship = kontextMat[["Mesenchymal__Macrophages__tissue"]]
survRelationship = ifelse(survRelationship > median(survRelationship), "Localised", "Dispersed")
    
# Plotting Kaplan-Meier curve
survfit2(kerenSurv ~ survRelationship) |>
    ggsurvfit() +
    add_confidence_interval() +
    add_pvalue() +
    ggtitle("Mesenchymal__Macrophages__tissue")
```


# Cross Validation

```{r}
survivalData <- kerenSCE %>%
  colData() %>%
  as.data.frame() %>%
  select(c("imageID", "Survival_days_capped")) %>%
  distinct()

modellingData <- listImageModelsCVFormat(stateChanges,
                                         values_from = "tValue",
                                         removeColsThresh = 0.2,
                                         missingReplacement = 0)
for(modelling in modellingData) {
  modelling <- modelling %>%
    left_join(survivalData, by = "imageID")
}
```

# LisaClust
```{r lisaClust}
set.seed(101)
data(kerenSCE)

regions <- lisaClust::lisaClust(kerenSCE, k = 5)

lisaClustOutput <- getMarkerMeans(regions,
                              survivalData = "Survival_days_capped")

```

# References

Keren, L., Bosse, M., Marquez, D., Angoshtari, R., Jain, S., Varma, S., Yang, S. R., Kurian, A., Van Valen, D., West, R., Bendall, S. C., & Angelo, M. (2018). A Structured Tumor-Immune Microenvironment in Triple Negative Breast Cancer Revealed by Multiplexed Ion Beam Imaging. Cell, 174(6), 1373-1387.e1319. ([DOI](https://doi.org/10.1016/j.cell.2018.08.039))



# sessionInfo
```{r}
sessionInfo()
```
